Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-24 manuscript/manuscript.tex
Shorten methods section and move to appendix

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from EVA were converted into into OpenVA format \citep{liOpenVAToolkitVerbal2023}, and fed into each model to produce COD assignments as WHOVA 2016 codes. All model outputs (ICD-10 for GPT-3.5/4 and WHOVA-2016 for InSilicoVA/InterVA-5) were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians. See Appendix \ref{app-methods-model} for additional details regarding input parameters and code conversions for each model.

@@

The performance of four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, were evaluated with metrics on the population and individual level by comparing their CGHR-10 COD outputs to 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{methods-eval-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{methods-eval-pccc}) \citep{murrayRobustMetricsAssessing2011}. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model by three age groups, and further into age ranges and COD for each age group.

The CSMF accuracy and PCCC metrics were calculated and compared for each model overall and by age group, followed by age ranges and COD for each age group as model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}. Metrics were calculated overall for three age groups according to the CGHR-10 codes: adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days). For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. Metrics were also calculated by CODs defined by CGHR-10 codes, which include 19, 10, and 7 CODs for adult, child, and neonatal records respectively.

@@

\subsubsection{Cause Specific Mortality Fraction (CSMF) Accuracy}\label{methods-eval-csmfa}

@@

\subsubsection{Partial Chance Corrected Concordance (PCCC)}\label{methods-eval-pccc}

>>>>

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from EVA were converted into into OpenVA format \citep{liOpenVAToolkitVerbal2023}, before being used as input for each model to produce COD assignments as WHO VA 2016 v1.5 \citep{worldhealthorganizationVerbalAutopsyStandards2016} codes. All model outputs (ICD-10 for GPT-3.5/4 and WHO VA 2016 for InSilicoVA/InterVA-5) were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians. See Appendix \ref{app-methods-model} for additional details regarding input parameters, output data, and code conversions for each model.

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal, by CGHR-10 COD codes, and across age ranges. For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

The performance of GPT-3.5, GPT-4, InSilicoVA, and InterVA-5 models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records to physician COD assignments. Section \ref{app-methods-csmfa} describes CSMF accuracy in detail for evaluating models on the population level,  Section \ref{app-methods-pccc} describes PCCC for evaluating models on the individual level. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model overall and by three age groups (adult, child, and neonatal), then further into age ranges and COD for each age group.

\subsubsection{Cause Specific Mortality Fraction (CSMF) Accuracy}\label{app-methods-csmfa}

@@

\subsubsection{Partial Chance Corrected Concordance (PCCC)}\label{app-methods-pccc}
